---
title: "Introduction to icd10us"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to icd10us}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(icd10us)
library(dplyr)
```

## Overview

The `icd10us` package provides comprehensive data and tools for working with ICD-10 codes in the United States healthcare context. It includes:

- **ICD-10-CM diagnosis codes** with metadata
- **ICD-10-PCS procedure codes** with metadata  
- Utility functions for transforming, filtering, and grouping codes

All datasets in this package are in the **public domain**, sourced from:

- Centers for Medicare and Medicaid Services (CMS)
- Centers for Disease Control and Prevention (CDC)
- Healthcare Cost and Utilization Project (HCUP) from the Agency for Healthcare Research and Quality (AHRQ)

## ICD-10-CM Diagnosis Codes

### The icd10cm Dataset

The `icd10cm` dataset contains all official ICD-10-CM diagnosis codes used in the United States:

```{r}
# View the structure of the dataset
str(icd10cm)

# See first few rows
head(icd10cm)
```

Each row represents one ICD-10-CM code with the following information:

- **icd10cm_code**: The code itself (without dot)
- **icd10cm_short_description**: Brief description
- **icd10cm_long_description**: Full description
- **valid_billing_code**: Whether the code can be used for billing (1=Yes, 0=No)
- **poa_exempt**: Present on Admission (POA) exemption status
- **chronic_indicator**: Classification as chronic (C) or acute (A) condition

### Filtering Codes

You can filter the dataset using standard `dplyr` functions:

```{r}
# Find all valid billing codes for diabetes
diabetes_codes <- icd10cm %>%
  filter(
    icd10cm_code >= "E08" & icd10cm_code <= "E13",
    valid_billing_code == 1
  )

head(diabetes_codes)
```

```{r}
# Find all chronic conditions
chronic_conditions <- icd10cm %>%
  filter(chronic_indicator == "C")

# Count chronic vs acute conditions
icd10cm %>%
  filter(!is.na(chronic_indicator)) %>%
  count(chronic_indicator)
```

### Chapter Organization

The `icd10cm_chapters` dataset organizes codes into 22 chapters:

```{r}
# View all chapters
icd10cm_chapters

# Find mental health chapter
icd10cm_chapters %>%
  filter(chapter_abbr == "Mental")
```

## Working with Code Ranges

### Getting Codes Between Two Codes

The `codes_between()` function retrieves all codes in a range:

```{r}
# Get all depression codes (F32-F33)
depression_codes <- codes_between("F32", "F33")

# View billable codes only
depression_codes %>%
  filter(valid_billing_code == 1) %>%
  select(icd10cm_code, icd10cm_short_description)
```

By default, `codes_between()` expands the end code to include all its children:

```{r}
# With expansion (default)
codes_with_expansion <- codes_between("F32", "F33", expand_end_code = TRUE)
nrow(codes_with_expansion)

# Without expansion
codes_without_expansion <- codes_between("F32", "F33", expand_end_code = FALSE)
nrow(codes_without_expansion)
```

### Expanding a Code

The `expand_code()` function shows all child (more specific) codes:

```{r}
# Expand F32 to see all subtypes
f32_codes <- expand_code("F32")

# Show billable codes
f32_codes %>%
  filter(valid_billing_code == 1) %>%
  select(icd10cm_code, icd10cm_short_description)
```

### Finding Header Codes

The `header_code()` function finds the parent/header code for a specific code:

```{r}
# Get header for a specific code
header_code("F320")

# Get header for an already-header code
header_code("F32")
```

## Code Validation and Formatting

### Validating Code Format

The `is_valid_icd10cm()` function checks if a string matches ICD-10-CM format:

```{r}
# Valid ICD-10-CM codes
is_valid_icd10cm("F320")    # TRUE
is_valid_icd10cm("E119")    # TRUE
is_valid_icd10cm("F32")     # TRUE

# Invalid codes
is_valid_icd10cm("296")     # FALSE - ICD-9 format
is_valid_icd10cm("F")       # FALSE - too short
is_valid_icd10cm("ABCD")    # FALSE - no numbers
```

Note: This only validates the **format**, not whether the code actually exists in the official code set.

### Adding and Removing Dots

ICD-10 codes can be formatted with or without decimal points:

```{r}
# Add dots
add_dot("F320")     # Returns "F32.0"
add_dot("E1165")    # Returns "E11.65"
add_dot("F32")      # Returns "F32" (3-char codes don't need dots)

# Remove dots
remove_dot("F32.0")   # Returns "F320"
remove_dot("E11.65")  # Returns "E1165"
remove_dot("F32")     # Returns "F32" (no dot to remove)
```

## ICD-10-PCS Procedure Codes

### The icd10pcs Dataset

The `icd10pcs` dataset contains all official ICD-10-PCS procedure codes:

```{r}
# View the structure
str(icd10pcs)

# See first few rows
head(icd10pcs)
```

Each procedure code includes a classification from HCUP:

- **Minor Diagnostic**: Non-operating room diagnostic procedures
- **Minor Therapeutic**: Non-operating room therapeutic procedures  
- **Major Diagnostic**: Operating room diagnostic procedures
- **Major Therapeutic**: Operating room therapeutic procedures

```{r}
# Count procedures by class
icd10pcs %>%
  filter(!is.na(procedure_class)) %>%
  count(procedure_class, sort = TRUE)
```

### Procedure Sections

The `icd10pcs_sections` dataset shows the top-level organization:

```{r}
# View all sections
icd10pcs_sections

# Medical and Surgical is the largest section
icd10pcs %>%
  filter(stringr::str_starts(icd10pcs_code, "0")) %>%
  nrow()
```

## Generating Sample Data

The `sample_icd10cm()` function generates synthetic diagnosis data for testing:

```{r}
# Generate 100 records for 10 patients
sample_data <- sample_icd10cm(n = 100, p = 10)
head(sample_data)

# Summary of the sample
sample_data %>%
  summarize(
    n_records = n(),
    n_patients = n_distinct(patient_id),
    n_unique_codes = n_distinct(icd10cm_code),
    date_range = paste(min(code_date), "to", max(code_date))
  )
```

You can also sample from specific code ranges:

```{r}
# Sample only mental health codes
mental_health_sample <- sample_icd10cm(
  n = 50, 
  p = 10,
  start_code = "F00",
  end_code = "F99"
)

# Verify all codes are F codes
mental_health_sample %>%
  pull(icd10cm_code) %>%
  head()
```

## Common Use Cases

### Use Case 1: Identifying Chronic Conditions

```{r}
# Create a function to identify chronic conditions in patient data
identify_chronic <- function(patient_codes) {
  patient_codes %>%
    left_join(
      icd10cm %>% select(icd10cm_code, chronic_indicator),
      by = "icd10cm_code"
    ) %>%
    filter(chronic_indicator == "C")
}

# Example with sample data
sample_patients <- sample_icd10cm(n = 50, p = 5)
chronic_diagnoses <- identify_chronic(sample_patients)

chronic_diagnoses %>%
  count(patient_id, sort = TRUE)
```

### Use Case 2: Grouping by Chapter

```{r}
# Assign chapter to diagnosis codes
assign_chapter <- function(codes_df) {
  # Create a helper function to find chapter
  find_chapter <- function(code) {
    icd10cm_chapters %>%
      filter(code_start <= code & code_end >= code) %>%
      pull(chapter_desc) %>%
      first()
  }
  
  codes_df %>%
    rowwise() %>%
    mutate(chapter = find_chapter(icd10cm_code)) %>%
    ungroup()
}

# Example
sample_with_chapters <- assign_chapter(sample_icd10cm(n = 30, p = 5))

sample_with_chapters %>%
  count(chapter, sort = TRUE)
```

### Use Case 3: Validating External Data

```{r}
# Suppose you have codes from an external source
external_codes <- c("F320", "E119", "Z99", "296", "INVALID")

# Validate format
validation_results <- tibble(
  code = external_codes,
  valid_format = is_valid_icd10cm(code),
  exists_in_codeset = code %in% icd10cm$icd10cm_code
)

validation_results
```

## Data Provenance and Public Domain Status

All datasets in this package are derived from **public domain** sources:

### ICD-10-CM Data Sources
- **Codes and Descriptions**: CMS via CDC/NCHS
- **POA Exemptions**: CDC/NCHS  
- **Chronic Condition Indicators**: AHRQ HCUP

### ICD-10-PCS Data Sources
- **Codes and Descriptions**: CMS
- **Procedure Classifications**: AHRQ HCUP

These data are freely available and can be used without restriction for research, clinical, and commercial purposes.

## Getting Help

For more information:

- Package documentation: `?icd10us`
- Function help: `?codes_between`, `?expand_code`, etc.
- GitHub repository: https://github.com/vcastro/icd10us
- Report issues: https://github.com/vcastro/icd10us/issues

## Session Info

```{r}
sessionInfo()
```
